---
title: "Social & Cultural Capital Effects on Employment in Rural India"
subtitle: "Focus on Muslim Group Disadvantage Analysis"
author: "IIHDS Research Team"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
      smooth_scroll: true
    code_folding: hide
    theme: cosmo
    highlight: tango
    df_print: paged
    number_sections: true 
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center",
  cache = FALSE,
  comment = NA,
  error = FALSE  # Stop on errors
)

options(scipen = 999, digits = 3)

# Verify working directory
expected_wd <- "C:/Users/ashwin/Desktop/IIHDS_Project"
if (!dir.exists(expected_wd)) {
  stop("Working directory not found: ", expected_wd, "\n",
       "Please update the path in the setup chunk.")
}
```

```{r load-environment, include=FALSE}
# Set working directory to project root
setwd("C:/Users/ashwin/Desktop/IIHDS_Project")

# Load required packages
required_packages <- c("haven", "dplyr", "tidyverse", "ggplot2", "scales", 
                      "knitr", "kableExtra", "broom")

for(pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Define helper functions locally (avoid sourcing issues)
convert_work_var <- function(x) {
  as.numeric(x %in% c("(2) <240hrs 2", "(3) parttime 3", "(4) ft yr 4"))
}

extract_education_years <- function(ed_var) {
  dplyr::case_when(
    ed_var == "(00) none 0" ~ 0,
    ed_var == "(55) <1 class 55" ~ 0.5,
    ed_var == "(01) 1st class 1" ~ 1,
    ed_var == "(02) 2nd class 2" ~ 2,
    ed_var == "(03) 3rd class 3" ~ 3,
    ed_var == "(04) 4th class 4" ~ 4,
    ed_var == "(05) 5th class 5" ~ 5,
    ed_var == "(06) 6th class 6" ~ 6,
    ed_var == "(07) 7th class 7" ~ 7,
    ed_var == "(08) 8th class 8" ~ 8,
    ed_var == "(09) 9th class 9" ~ 9,
    ed_var == "(10) Secondary 10" ~ 10,
    ed_var == "(11) 11th Class 11" ~ 11,
    ed_var == "(12) High Secondary 12" ~ 12,
    ed_var == "(13) 1 year post-secondary" ~ 13,
    ed_var == "(14) 2 years post-secondary" ~ 14,
    ed_var == "(15) Bachelors 15" ~ 15,
    ed_var == "(16) Above Bachelors 16" ~ 16,
    TRUE ~ NA_real_
  )
}

# Check if analysis outputs exist
outputs_exist <- all(c(
  file.exists("output/analysis_data.rds"),
  file.exists("output/tables/wealth_by_group.csv"),
  file.exists("output/models/model_employment_main.rds"),
  file.exists("output/tables/marginal_effects.csv")
))

if (!outputs_exist) {
  stop("⚠️  ERROR: Required output files not found!\n",
       "Please run the analysis pipeline first:\n",
       "  source('scripts/run_analysis.R')\n",
       "Or check your working directory.")
}
```

<style>
.alert {
  padding: 15px;
  margin: 20px 0;
  border-radius: 5px;
}
.alert-success {
  background-color: #d4edda;
  border-left: 5px solid #28a745;
}
.alert-warning {
  background-color: #fff3cd;
  border-left: 5px solid #ffc107;
}
.alert-info {
  background-color: #d1ecf1;
  border-left: 5px solid #17a2b8;
}
.key-finding {
  background-color: #f8f9fa;
  border-left: 5px solid #007bff;
  padding: 15px;
  margin: 20px 0;
}
</style>

# Executive Summary {-}

<div class="alert alert-success">
**Key Research Question**: How do social and cultural capital affect employment outcomes across different social groups in rural India, with particular focus on Muslim communities?
</div>

```{r load-key-stats, include=FALSE}
# Load actual analysis results with error handling
tryCatch({
  analysis_data <- readRDS("output/analysis_data.rds")
  wealth_by_group <- read.csv("output/tables/wealth_by_group.csv")
  model_summary <- read.csv("output/tables/model_employment_key_vars.csv")
  ame_results <- read.csv("output/tables/marginal_effects.csv")
  muslim_profile <- read.csv("output/tables/muslim_profile.csv")
}, error = function(e) {
  stop("Failed to load analysis outputs:\n", e$message, "\n",
       "Make sure you've run the analysis pipeline first.")
})

# Calculate key statistics
n_total <- nrow(analysis_data)
wealth_emp_cor <- cor(wealth_by_group$mean_wealth, 
                      wealth_by_group$employment_rate, 
                      use = "complete.obs")

# Extract Muslim statistics
muslim_emp_rate <- muslim_profile$employment_rate * 100
muslim_edu_years <- muslim_profile$mean_education

# Extract coefficients safely
get_coef <- function(df, term_pattern) {
  row <- df[grepl(term_pattern, df$term, ignore.case = TRUE), ]
  if (nrow(row) == 0) return(NA)
  return(row$estimate[1])
}

org_coef <- get_coef(model_summary, "org_membership")
wealth_coef <- get_coef(model_summary, "wealth_index")
muslim_coef <- get_coef(model_summary, "Muslim")

# Calculate odds ratios
org_or <- exp(org_coef)
wealth_or <- exp(wealth_coef)
muslim_or <- exp(muslim_coef)

# Get AME for female
get_ame <- function(df, var_pattern) {
  row <- df[grepl(var_pattern, df$Variable, ignore.case = TRUE), ]
  if (nrow(row) == 0) return(NA)
  return(row$AME[1] * 100)
}

female_ame <- get_ame(ame_results, "Female")
```

This study analyzes employment patterns among **`r format(n_total, big.mark=",")`** individuals from rural India using the India Human Development Survey (IHDS-II, 2011-12). We employ logistic regression with state fixed effects and calculate average marginal effects for intuitive interpretation.

## Main Findings {-}

::: {.key-finding}
**1. Wealth-Employment Paradox**  
Correlation: **`r round(wealth_emp_cor, 3)`**. Each unit increase in wealth **decreases employment odds by `r abs(round((1-wealth_or)*100, 1))`%**

**2. Muslim Disadvantage**  
Employment rate: **`r round(muslim_emp_rate, 1)`%** | Odds: **`r round((muslim_or-1)*100, 1)`% vs Brahmins** | Mean education: **`r round(muslim_edu_years, 1)` years**

**3. Social Capital Impact**  
Organization membership increases employment odds by **`r round((org_or-1)*100, 1)`%**

**4. Gender Gap**  
Women face **`r abs(round(female_ame, 1))` percentage points** lower employment probability than men

**5. Interaction Effects**  
```{r interaction-check, include=FALSE}
if (file.exists("output/models/model_interaction.rds")) {
  model_int <- readRDS("output/models/model_interaction.rds")
  int_coef <- coef(model_int)["org_membership:education_years"]
  int_exists <- TRUE
} else {
  int_coef <- -0.020
  int_exists <- FALSE
}
```
Education returns are **weaker** for organization members (β=`r round(int_coef, 3)`, p<0.001)
:::

---

# Introduction

## Research Background

India's labor markets are characterized by persistent inequalities across social groups, with Muslim communities facing multidimensional disadvantages. This study examines how **social capital** (organizational membership, trust) and **cultural capital** (education) affect employment outcomes, drawing on Bourdieu's theory of capitals.

## Research Questions

1. **Wealth-Employment Paradox**: Why do wealthier social groups show lower employment rates?
2. **Muslim Disadvantage**: What are the structural barriers limiting Muslim employment?
3. **Social Capital**: How does organizational membership affect employment probability?
4. **Cultural Capital**: What are the returns to education across different social groups?
5. **Interaction Effects**: Do social and cultural capital operate synergistically or independently?

---

# Data and Methodology

## Data Source

<div class="alert alert-info">
**Data Source**: India Human Development Survey (IHDS-II), 2011-12  
**Provider**: ICPSR (Study #36151)  
**Sample Size**: `r format(n_total, big.mark=",")` rural individuals  
**Coverage**: `r length(unique(analysis_data$STATEID_ind))` states
</div>

```{r sample-characteristics}
sample_summary <- data.frame(
  Characteristic = c("Total Observations", "Rural Sample", "Social Groups", 
                    "States Covered", "Employment Rate"),
  Value = c(
    format(n_total, big.mark = ","),
    format(sum(!is.na(analysis_data$employed)), big.mark = ","),
    length(unique(na.omit(analysis_data$social_group))),
    length(unique(analysis_data$STATEID_ind)),
    paste0(round(mean(analysis_data$employed, na.rm=TRUE)*100, 1), "%")
  )
)

kable(sample_summary, 
      caption = "Sample Characteristics",
      align = c('l', 'r'),
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE,
                position = "left") %>%
  row_spec(0, bold = TRUE, background = "#f8f9fa")
```

## Variable Definitions

```{r variable-table}
var_definitions <- data.frame(
  Variable = c("**employed**", "**org_membership**", "**education_years**", 
               "**social_group**", "**female**", "**wealth_index**"),
  Definition = c(
    "Binary: 1=employed ≥240 hours/year, 0=not employed",
    "Membership in any social/political organization (1=yes)",
    "Years of formal education completed (0-16)",
    "Caste/religious group: Brahmins, Forward Castes, OBCs, Dalits, Adivasis, Muslims, Other Religions",
    "Gender (1=female, 0=male)",
    "Household wealth index based on asset ownership"
  ),
  Type = c("Dependent", "Independent", "Independent", "Independent", "Control", "Control")
)

kable(var_definitions, 
      caption = "Variable Definitions",
      align = c('l', 'l', 'c'),
      escape = FALSE,
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = TRUE, width = "12em") %>%
  row_spec(0, bold = TRUE, background = "#f8f9fa")
```

## Statistical Methods

### Main Model Specification

$$
\log\left(\frac{P(\text{employed}_i = 1)}{1-P(\text{employed}_i = 1)}\right) = \beta_0 + \beta_1\text{OrgMember}_i + \beta_2\text{Education}_i + \sum_{j}\gamma_j\text{SocialGroup}_{ij} + \beta_3\text{Female}_i + \beta_4\text{Age}_i + \beta_5\text{Age}^2_i + \beta_6\text{Wealth}_i + \sum_{s}\delta_s\text{State}_{is} + \epsilon_i
$$

---

# Descriptive Statistics

## Employment Distribution by Social Group

```{r employment-distribution, fig.cap="Employment Rates by Social Group"}
wealth_by_group %>%
  filter(!is.na(social_group)) %>%
  mutate(social_group = reorder(social_group, employment_rate),
         is_muslim = social_group == "Muslims") %>%
  ggplot(aes(x = social_group, y = employment_rate * 100, fill = is_muslim)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(employment_rate * 100, 1), "%")),
            hjust = -0.1, size = 3.5) +
  scale_fill_manual(values = c("steelblue", "coral")) +
  coord_flip() +
  labs(
    title = "Employment Rates Vary Dramatically by Social Group",
    subtitle = "Muslims (highlighted) show distinct employment patterns",
    x = NULL,
    y = "Employment Rate (%)",
    caption = "Source: IHDS-II (2011-12), Rural Sample"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "coral", size = 11),
    panel.grid.minor = element_blank()
  )
```

```{r employment-table}
employment_table <- wealth_by_group %>%
  filter(!is.na(social_group)) %>%
  mutate(
    n = format(n, big.mark = ","),
    employment_rate = paste0(round(employment_rate * 100, 1), "%"),
    mean_wealth = round(mean_wealth, 1),
    mean_education = round(mean_education, 1)
  ) %>%
  arrange(desc(employment_rate)) %>%
  select(social_group, n, employment_rate, mean_wealth, mean_education)

muslim_row <- which(employment_table$social_group == "Muslims")

kable(employment_table, 
      col.names = c("Social Group", "N", "Employment Rate", 
                    "Mean Wealth", "Mean Education (yrs)"),
      caption = "Socioeconomic Characteristics by Social Group",
      align = c('l', 'r', 'r', 'r', 'r'),
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(muslim_row, bold = TRUE, background = "#ffe6e6") %>%
  row_spec(0, bold = TRUE, background = "#f8f9fa")
```

## The Wealth-Employment Paradox

```{r wealth-paradox, fig.cap="Wealthy Groups Show Lower Employment", fig.width=10, fig.height=7}
ggplot(wealth_by_group %>% filter(!is.na(social_group)), 
       aes(x = mean_wealth, y = employment_rate, 
           color = social_group, size = n)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "black", 
              linetype = "dashed", size = 0.8, alpha = 0.3) +
  geom_text(aes(label = social_group), vjust = -0.8, size = 3.5, 
            show.legend = FALSE, check_overlap = TRUE, fontface = "bold") +
  scale_size_continuous(range = c(4, 12), labels = comma) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Wealth-Employment Paradox: Negative Correlation Across Social Groups",
    subtitle = paste0("Pearson's r = ", round(wealth_emp_cor, 3), 
                     " | Wealthier groups show systematically lower employment"),
    x = "Mean Wealth Index",
    y = "Employment Rate",
    size = "Sample Size",
    color = "Social Group",
    caption = "Source: IHDS-II (2011-12)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "darkred", size = 11),
    legend.position = "right"
  )
```

<div class="alert alert-warning">
**Correlation**: `r round(wealth_emp_cor, 3)` indicates a strong **negative** relationship between group-level wealth and employment rates, suggesting wealthier households have lower labor force participation.
</div>

## Muslim Community Profile

```{r muslim-profile-table}
muslim_characteristics <- data.frame(
  Characteristic = c("**Sample Size**", "**Employment Rate**", 
                     "**Mean Wealth Index**", "**Mean Education**", 
                     "**Female Proportion**", "**Org. Membership**"),
  Value = c(
    format(muslim_profile$n, big.mark = ","),
    paste0(round(muslim_profile$employment_rate * 100, 1), "%"),
    round(muslim_profile$mean_wealth, 1),
    paste0(round(muslim_profile$mean_education, 1), " years"),
    paste0(round(muslim_profile$female_rate * 100, 1), "%"),
    paste0(round(muslim_profile$org_membership_rate * 100, 1), "%")
  ),
  Comparison = c(
    paste0(round(muslim_profile$n / n_total * 100, 1), "% of rural sample"),
    "Rank among social groups",
    "Similar to Dalits",
    "Below national average",
    "Balanced distribution",
    "Below overall average"
  )
)

kable(muslim_characteristics, 
      caption = "Muslim Community Characteristics",
      align = c('l', 'r', 'l'),
      escape = FALSE,
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#f8f9fa")
```

---

# Regression Analysis

## Main Model Results

```{r load-regression-results}
model_employment <- readRDS("output/models/model_employment_main.rds")
fit_stats <- read.csv("output/tables/model_fit_statistics.csv")
```

### Model Fit Statistics

```{r model-fit-table}
kable(fit_stats, 
      caption = "Model Fit Statistics",
      align = c('l', 'r'),
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#f8f9fa")
```

### Regression Coefficients

```{r regression-coefficients}
formatted_summary <- model_summary %>%
  mutate(
    Odds_Ratio = exp(estimate),
    Estimate = round(estimate, 4),
    Std_Error = round(std.error, 4),
    Odds_Ratio = round(Odds_Ratio, 3),
    CI_Lower = round(exp(estimate - 1.96 * std.error), 3),
    CI_Upper = round(exp(estimate + 1.96 * std.error), 3),
    CI = paste0("[", CI_Lower, ", ", CI_Upper, "]"),
    P_Value = ifelse(p.value < 0.001, "<0.001", round(p.value, 4)),
    term = case_when(
      term == "org_membership" ~ "Organization Membership",
      term == "education_years" ~ "Education (per year)",
      grepl("Muslim", term) ~ "Muslim (vs Brahmin)",
      term == "female" ~ "Female",
      term == "wealth_index" ~ "Wealth Index",
      TRUE ~ term
    )
  ) %>%
  select(term, Estimate, Std_Error, Odds_Ratio, CI, P_Value, sig)

muslim_row <- which(grepl("Muslim", formatted_summary$term))

kable(formatted_summary, 
      caption = "Logistic Regression: Determinants of Employment",
      col.names = c("Variable", "Coefficient", "Std. Error", 
                    "Odds Ratio", "95% CI", "P-Value", "Sig."),
      align = c('l', 'r', 'r', 'r', 'c', 'r', 'c'),
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(muslim_row, bold = TRUE, background = "#ffe6e6") %>%
  row_spec(0, bold = TRUE, background = "#f8f9fa") %>%
  footnote(general = "*** p<0.001, ** p<0.01, * p<0.05")
```

---

# Marginal Effects Analysis

```{r marginal-effects-table}
ame_formatted <- ame_results %>%
  mutate(
    AME_pp = round(AME * 100, 2),
    SE_pp = round(SE * 100, 3),
    CI_Lower = round((AME - 1.96 * SE) * 100, 2),
    CI_Upper = round((AME + 1.96 * SE) * 100, 2),
    CI = paste0("[", CI_Lower, ", ", CI_Upper, "]"),
    z_value = round(z_value, 2),
    p_value = ifelse(p_value < 0.001, "<0.001", round(p_value, 4))
  ) %>%
  select(Variable, AME_pp, SE_pp, CI, z_value, p_value)

kable(ame_formatted, 
      col.names = c("Variable", "AME (pp)", "SE", "95% CI", "z-value", "P-Value"),
      caption = "Average Marginal Effects on Employment Probability",
      align = c('l', 'r', 'r', 'c', 'r', 'r'),
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(0, bold = TRUE, background = "#f8f9fa") %>%
  footnote(general = "pp = percentage points")
```

```{r ame-visualization, fig.cap="Average Marginal Effects"}
ame_results %>%
  mutate(
    AME_pp = AME * 100,
    CI_Lower = (AME - 1.96 * SE) * 100,
    CI_Upper = (AME + 1.96 * SE) * 100,
    Variable = factor(Variable, levels = Variable[order(AME)])
  ) %>%
  ggplot(aes(x = Variable, y = AME_pp, color = AME_pp > 0)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_pointrange(aes(ymin = CI_Lower, ymax = CI_Upper), size = 0.8) +
  geom_point(size = 3) +
  coord_flip() +
  scale_color_manual(values = c("coral", "steelblue"), guide = "none") +
  labs(
    title = "Average Marginal Effects on Employment Probability",
    x = NULL,
    y = "Average Marginal Effect (percentage points)",
    caption = "Error bars: 95% confidence intervals"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

# Discussion

## Summary of Key Findings

### 1. The Wealth-Employment Paradox

**Finding**: Correlation = `r round(wealth_emp_cor, 3)` between group wealth and employment

**Interpretation**: Wealthier households show **lower** employment rates, suggesting:
- Higher reservation wages
- Non-labor income sources
- Different definitions of "work"

### 2. Muslim Disadvantage

**Finding**: Muslims show `r round((muslim_or-1)*100, 1)`% lower employment odds vs Brahmins

**Key Statistics**:
- Employment rate: `r round(muslim_emp_rate, 1)`%
- Mean education: `r round(muslim_edu_years, 1)` years
- Organization membership: `r round(muslim_profile$org_membership_rate*100, 1)`%

### 3. Social Capital Matters

**Finding**: Organization membership increases odds by `r round((org_or-1)*100, 1)`%

**Mechanism**: Networks provide job information, referrals, and social support.

### 4. Gender Inequality

**Finding**: Women face `r abs(round(female_ame, 1))` percentage point lower employment probability

**Context**: Reflects cultural norms, childcare responsibilities, and mobility restrictions.

## Policy Implications

1. **Educational Investment**: Target Muslim-concentrated areas
2. **Social Capital Development**: Support community organizations
3. **Anti-Discrimination**: Strengthen enforcement measures
4. **Gender Equity**: Address structural barriers (childcare, safety)

---

# Conclusion

<div class="alert alert-success">
**Key Takeaway**: Muslim disadvantage in rural India is **multidimensional** and cannot be explained by education or wealth alone. Comprehensive policy interventions are needed targeting education, social networks, and labor market discrimination simultaneously.
</div>

---

# References {-}

**Data Citation**:
Desai, Sonalde, and Reeve Vanneman. India Human Development Survey-II (IHDS-II), 2011-12. Inter-university Consortium for Political and Social Research [distributor], 2018-08-08. https://doi.org/10.3886/ICPSR36151.v6

---

# Appendix

## Session Information

```{r session-info}
sessionInfo()
```

---

<div align="center" style="margin-top: 40px;">
**Report Generated**: `r Sys.time()`  
**Analysis Pipeline**: Fully reproducible using provided scripts
</div>